---
title: "1.1 data wrangling"
author: "TG"
date: "2023-12-10"
output:
   prettydoc::html_pretty:
    theme: cayman
    highlight: github
---


```{r message=TRUE, warning=TRUE, include=FALSE}

library(tidyverse)
library(ggplot2)
library(stringr)
library(Hmisc)
library(rfishbase)
library(lubridate)

select <- dplyr::select
rename <-dplyr::rename
```

# Coral data wrangling

load data

```{r}
# old data until 2021

load("G:/My Drive/PHD/Ecocamp-2022/data/until 2021/coral_comm.RData")
load("G:/My Drive/PHD/Ecocamp-2022/data/until 2021/coral_cover.RData")

# knoll medata (don't change along the years)

load("G:/My Drive/PHD/Ecocamp-2022/data/until 2021/knoll_mdata.RData")

# data from 2022

cpc_2022 <- read_csv("data/2022/cpc 2022.csv")

```


Shaping `cpc_2022` to the format of `coral_comm`

```{r}


comm_22<-cpc_2022

# unify column names

colnames(comm_22)<-c("site","knoll","Category","species","points")

# some the number of point for each species

comm_22<-comm_22 %>% group_by(site,knoll,Category,species) %>% summarise("points" = sum(points))

# change to similar site names

comm_22<-comm_22 %>% mutate(site = case_when(site == "CA"~"Caves",
                                             site == "NR"~"NR",
                                             site == "KA"~"Katza",
                                             site == "PR"~"Princess",))


# correct the knoll number in princess 


comm_22$knoll<-ifelse(comm_22$site =="Princess" & comm_22$knoll =="21",22,comm_22$knoll)

comm_22$knoll<-ifelse(comm_22$site =="Princess" & comm_22$knoll =="10",23,comm_22$knoll)

# add new columns of year, storm and knoll id

comm_22<-comm_22 %>%
  mutate(year = rep(2022),.after = knoll) %>%
  mutate(storm = rep("After"),.after = year) %>% 
  mutate(knoll_id = paste(site,knoll,storm,sep = "_"),.after = storm)


# unify category names

a<-unique(comm_22$Category)
b<-unique(coral_comm$Category)
a[which(!a %in% b)]


comm_22$Category[comm_22$Category == "Soft Corals"]<- "Soft Coral"
comm_22$Category[comm_22$Category == "Coral"]<- "Stony Coral"
comm_22$Category[comm_22$Category == "Tunicate"]<- "Ascidia"
comm_22$Category[comm_22$Category == "Sea Anemone"]<- "Sea anemone"

# rock appear in the data but not in the coral community data, ill remove it after joining

rm("a","b")

# unify species names

a<-unique(comm_22$species)
b<-unique(coral_comm$species)
a[which(!a %in% b)]


coral_comm$species[coral_comm$species == "Dipsastrea"]<- "Dipsastraea"
coral_comm$species[coral_comm$species == "Paramontastrea"]<- "Paramontastraea"
coral_comm$species[coral_comm$species == "Montastrea"]<- "Montastraea"


comm_22$species[comm_22$species == "Stony Coral"]<- "Stony coral"
comm_22$species[comm_22$species == "Sea Anemone"]<- "Sea anemone"
comm_22$species[comm_22$species == "Oxypora"]<- "Oxyopora"
comm_22$species[comm_22$species == "Siderastraea"]<- "Siderastrea"
comm_22$species[comm_22$species == "Soft Coral"]<- "Soft coral"


# join with knoll meta data

knoll_mdata_join<-knoll_mdata %>% select(4:ncol(knoll_mdata))


comm_22<-left_join(comm_22,knoll_mdata_join)

# make sure columns are in the same order

comm_22<-comm_22 %>% select(colnames(coral_comm))

# united coral comm

coral_community_updated_2022<-bind_rows(coral_comm,comm_22)

#filter the rock which appear only in 2022 data

coral_community_updated_2022<-coral_community_updated_2022 %>% filter(Category != "Rock")

```


Shaping `cpc_2022` to the format of `coral_cover`



```{r}


comm_22$Category<-sub(" ", ".", comm_22$Category)

comm_22$Category[comm_22$Category=="Rock"]<-"Other"

a<- unique(comm_22$Category)
b<- colnames(coral_cover[6:17])

a[which(!a %in% b)]
b[which(!b %in% a)]

rm("a","b")

# calculate precentage

cover_22<-comm_22 %>% select(-species)
cover_22<-cover_22 %>% relocate(6:7,.after = ITM_Y)

cover_22<-cover_22 %>% group_by(across(1:18)) %>% summarise(points = sum(points))


cover_22<-cover_22 %>% group_by(across(1:17)) %>% mutate("knoll_sum" = sum(points)) %>% ungroup()


cover_22<-cover_22 %>% mutate("precentage" = points*100/knoll_sum)

cover_22 <-cover_22 %>% select(-points,-knoll_sum)

# convert to wide format

cover_22<-spread(cover_22,Category,precentage,fill = 0)

# match the column names before Union 

a<-colnames(cover_22)
b<-colnames(coral_cover)

a[which(!a %in% b)]
b[which(!b %in% a)]

cover_22<-cover_22 %>% rename(Site = site,
                              Year = year,
                              Knoll = knoll)

cover_22<-cover_22 %>% mutate(Algae = rep(0),
                              Corallimorpharia = rep(0),
                              Echinoderms = rep(0),
                              Mollusca = rep(0),
                              Zoantharia = rep(0))

cover_22<-cover_22 %>% select(colnames(coral_cover))


rm("a","b")


coral_cover_updated_2022<- bind_rows(coral_cover,cover_22)
```

### save coral data

```{r}

dir_2022<- "G:/My Drive/PHD/Ecocamp-2022/data/2022/clean coral data"
dir_all<- "G:/My Drive/PHD/Ecocamp-2022/data/all years"

# clean 2022 data

save(cover_22,file = paste0(dir_2022,sep = "/coral_cover_22.Rdata"))
save(comm_22,file = paste0(dir_2022,sep = "/coral_comm_22.Rdata"))


# all data united

save(coral_cover_updated_2022,
     file = paste0(dir_all,sep = "/coral_cover_updated_2022.Rdata"))

write.csv(coral_cover_updated_2022,file = paste0(dir_all,sep = "/coral_cover_updated_2022.csv"))

save(coral_community_updated_2022,
     file = paste0(dir_all,sep = "/coral_community_updated_2022.Rdata"))

write.csv(coral_community_updated_2022,
     file = paste0(dir_all,sep = "/coral_community_updated_2022.csv"))
```


# Fish data wrangling

load data

```{r}
# clean and orgenise data from 2015-2021

load("G:/My Drive/PHD/Ecocamp-2022/data/until 2021/fish_full_data.Rdata") # old data

# new fish data from fall 2022

EcoCamp_2022_raw <- read_csv("data/2022/fish ecocamp 2022.csv")

data_2022<-EcoCamp_2022_raw


```


```{r}

# change firsy/second observer to the observer name

data_2022$Observer <- ifelse(data_2022$Observer == "first", 
                            data_2022$`First Observer.x`,
                            data_2022$`Second Observer.x`)

# keep only relevant column

data_2022 <- data_2022 %>%
  rename(Site = Location, Abundance = Amount) %>%
  mutate(Date = ymd(Date),Year = year(Date)) %>% 
  select(Site,Knoll,Year,Date,Observer,Species, Abundance, Length,Confidence)

# convert column types 

data_2022$Site = as.character(data_2022$Site) 
data_2022$Year = as.character(data_2022$Year) 
data_2022$Knoll = as.character(data_2022$Knoll)
# change site name to NR

data_2022$Site[data_2022$Site == "Reserve"] = "NR"

# In cases where there is no confidance convert to 0 

data_2022$Confidence[is.na(data_2022$Confidence)]<-0

```


Insert correction to knolls number

knoll 21 in Princess should change to 22 after the storm (2020)
knoll 10 in the Princess should change to 23 after the storm (2020)

Those knoll was documented under the same knoll but the knoll was destroy in the storm and we actually surveyed different knoll!


```{r}

data_2022$Knoll<-ifelse(data_2022$Site =="Princess" & data_2022$Knoll =="21",22,data_2022$Knoll)

data_2022$Knoll<-ifelse(data_2022$Site =="Princess" & data_2022$Knoll =="10",23,data_2022$Knoll)

```


check species names to match fish full data and the updated name accurding to fishbase


```{r}
# list of distinct species

sp_2022<-data_2022 %>% distinct(Species)

# check which fish are valdiate accurding to fishbase

valid_name <-validate_names(sp_2022$Species)

sp_2022$valid<-ifelse(sp_2022$Species %in% valid_name,"valid","not valid")

# remove valid species and sp/spp observations

species_v<-sp_2022 %>% filter(sp_2022$valid == "not valid")
species_v<-species_v %>% filter(str_detect(species_v$Species,"spp")==F)
species_v<-species_v %>% filter(str_detect(species_v$Species,"sp.")==F)
 
# match to the new and accepted name

new_name<-species_v$Species %>% synonyms() %>% 
 select(provided_name = synonym, valid_name = Species, Comment = Status) %>% 
  filter(Comment == "synonym") %>% rename(Species = provided_name)

# join and replace the previos name

data_2022<-left_join(data_2022,new_name,by="Species")

data_2022$Species<-ifelse(!is.na(data_2022$valid_name),
                                data_2022$valid_name,
                                data_2022$Species)

data_2022<-data_2022[,1:9]

```

change species that are found in different distribution range to the local species

```{r}
data_2022$Species[data_2022$Species == "Dascyllus aruanus"]<-"Dascyllus abudafur"
data_2022$Species[data_2022$Species == "Macropharyngodon bipartitus"]<-"Macropharyngodon marisrubri"
data_2022$Species[data_2022$Species == "Aetobatus narinari"]<-"Aetobatus ocellatus"
data_2022$Species[data_2022$Species == "Amblyglyphidodon leucogaster"]<-"Amblyglyphidodon indicus"
data_2022$Species[data_2022$Species == "Parapriacanthus ransonneti"]<-"Parapriacanthus guentheri"
data_2022$Species[data_2022$Species == "Eviota sebreei"]<-"Eviota punyit"
data_2022$Species[data_2022$Species == "Eviota prasites"]<-"Eviota zebrina"
data_2022$Species[data_2022$Species == "Canthigaster coronata"]<-"Canthigaster cyanospilota"
data_2022$Species[data_2022$Species == "Hemigymnus fasciatus"]<-"Hemigymnus sexfasciatus"

```

check which species are new compared to the data collected up to 2021

```{r}

sp_2022<-data_2022 %>% distinct(Species)

sp_all<-fish_full_data%>% distinct(Species,.keep_all = T) %>% select(Species,Species_2015,Family,Status,a,b)

sp_2022$Species[which(!sp_2022$Species %in% sp_all$Species)]

```

* `Pycnochromis dimidiatus` - updated name, need to change in the full data
* `Stegastes lacrymatus` - updated name, need to change in the full data
* `Gymnothorax johnsoni`,`Gymnomuraena zebra` and `Caesio suevica` - new species in the data


change the updated names in fish_full_data

```{r}

fish_full_data$Species[fish_full_data$Species == "Chromis dimidiata"]<-"Pycnochromis dimidiatus"
fish_full_data$Species[fish_full_data$Species == "Plectroglyphidodon lacrymatus"]<-"Stegastes lacrymatus"



```

add family, coefficient and transient/reef associated status to the species


```{r}
sp_all<-fish_full_data%>% distinct(Species,.keep_all = T) %>% select(Species,Species_2015,Family,Status,a,b)

sp_2022<-left_join(sp_2022,sp_all)

```

Fill in the missing data for the 3 species

*I'll add the a and b later (when fishbase will work)*

```{r}
sp_2022$Species_2015<-ifelse(sp_2022$Species == "Gymnothorax johnsoni"|
                               sp_2022$Species == "Gymnomuraena zebra",
                             "Muraenidae",
                             sp_2022$Species_2015)



sp_2022$Family<-ifelse(sp_2022$Species == "Gymnothorax johnsoni"|
                         sp_2022$Species == "Gymnomuraena zebra",
                             "Muraenidae",
                             sp_2022$Family)


sp_2022$Status<-ifelse(sp_2022$Species == "Gymnothorax johnsoni"|
                         sp_2022$Species == "Gymnomuraena zebra",
                             "C",
                             sp_2022$Status)



sp_2022$Species_2015<-ifelse(sp_2022$Species == "Caesio suevica",
                             "Caesio spp.",
                             sp_2022$Species_2015)



sp_2022$Family<-ifelse(sp_2022$Species == "Caesio suevica",
                             "Caesionidae",
                             sp_2022$Family)


sp_2022$Status<-ifelse(sp_2022$Species == "Caesio suevica",
                             "T",
                             sp_2022$Status)




```



add a and b coefficents manually

```{r}
# from fishbase

sp_2022$a[sp_2022$Species=="Gymnomuraena zebra"]<- 0.00005	
sp_2022$b[sp_2022$Species=="Gymnomuraena zebra"]<-3.760   

# no coefficient in fishbase, used the one from the fishbase Bayesian model 

sp_2022$a[sp_2022$Species=="Gymnothorax johnsoni"]<- 0.00047	
sp_2022$b[sp_2022$Species=="Gymnothorax johnsoni"]<-3.28  


# no coefficient in fishbase, used the one from the fishbase Bayesian model 

sp_2022$a[sp_2022$Species=="Caesio suevica"]<- 0.01259	
sp_2022$b[sp_2022$Species=="Caesio suevica"]<- 3.09  
```


```{r}
data_2022<-left_join(data_2022,sp_2022)
```

Add `survey_id` and `year_month` columns

```{r}

data_2022<-data_2022 %>%  mutate("survey_id"= paste(Site,Knoll,Date,sep = "_"),.after = Date)

data_2022 <- data_2022 %>%  mutate("year_month"= paste(data_2022$Year,"b",sep=" "),.after = Date)

data_2022 <- data_2022 %>% relocate(9:11,.after = Family)


data_2022 <- data_2022 %>%  mutate("year_observer"= paste(Observer,Year,sep = "_"),.after = Confidence)
```

remove new surveyor data:


```{r}
unique(data_2022$Observer)

to_remove<-c("Yuval Goth","Inbar Dahan","Dana Kruger","Roi Holzman")

data_2022 <- data_2022 %>% filter(!Observer %in% to_remove)

rm("sp_2022","new_name","sp_all","species_v")
```


### add knoll metadata

```{r}
data_2022$storm <-rep("After")

data_2022$knoll_id<-paste(data_2022$Site,data_2022$Knoll,data_2022$storm,sep="_")

data_2022$Knoll<-as.numeric(data_2022$Knoll)

data_2022<-left_join(data_2022,knoll_mdata)


```

* need to add lat and lon for knoll NR_18_after,Princess_23_After & Princess_22_After


```{r}

coral_add <- cover_22

coral_add <- coral_add %>% mutate(coral_att = paste(Site,Knoll,Year,sep = "_"))


coral_add <- coral_add %>% select(coral_att,Algae,Soft.Coral,Stony.Coral)

coral_add <- coral_add %>% mutate(coral_cover = Soft.Coral+Stony.Coral)


coral_add <- coral_add %>% rename(Soft_Coral = Soft.Coral,
                                  Stony_Coral = Stony.Coral)

data_2022<- data_2022 %>% mutate(coral_att = paste(Site,Knoll,Year,sep = "_"))


data_2022<-left_join(data_2022,coral_add)

data_2022<-data_2022 %>% select(colnames(fish_full_data))

data_2022$Knoll<-as.character(data_2022$Knoll)

fish_full_data_updated_2022<-bind_rows(fish_full_data,data_2022)


```


# save fish data


```{r}
save(fish_full_data_updated_2022,
     file = paste0(dir_all,sep = "/fish_full_data_updated_2022.Rdata"))

save(knoll_mdata,
     file = paste0(dir_all,sep = "/knoll_mdata_updated_2022.Rdata"))
```


















# for assaf


```{r}

old_data<-fish_full_data %>% select(colnames(data_2022))

assaf_fish<-bind_rows(old_data,data_2022)

one_observer<-list()
loop<-1
set.seed(1)

for (i in unique(assaf_fish$survey_id)) {
  
  data <- assaf_fish %>% filter(survey_id == i)
  
  obs_amount <- length(unique(data$Observer))
  obs_name <- unique(data$Observer)
  chosen_obs <- sample(obs_name,1)  
 
  filter_data<-data %>% filter(Observer == chosen_obs)
  one_observer[[loop]]<-filter_data
  loop<-loop+1
  
  
}

assaf_fish<- bind_rows(one_observer)

rm(data,filter_data)
```




```{r}
assaf_fish$Year<-as.numeric(assaf_fish$Year)


assaf_fish <- assaf_fish %>%
  filter(Year > 2017,Confidence <1) %>%
  filter(! Knoll %in% c("Tamar",
                    "South bridge reserve",
                    "south bridge reserve",
                    "North bridge reserve",
                    "north bridge reserve"))

assaf_fish_per_knoll<-assaf_fish %>%
  group_by(Site,Knoll,year_month,Species) %>%
  summarise("Abundance" = sum(Abundance))


assaf_fish_per_site<-assaf_fish %>%
  group_by(Site,year_month,Species) %>%
  summarise("Abundance" = sum(Abundance))

assaf_fish_per_year <-assaf_fish %>%
  group_by(year_month,Species) %>%
  summarise("Abundance" = sum(Abundance))


#write.csv(assaf_fish,"assaf_fish.csv")
#write.csv(assaf_fish_per_knoll,"assaf_fish_per_knoll.csv")
#write.csv(assaf_fish_per_site,"assaf_fish_per_site.csv")
#write.csv(assaf_fish_per_year,"assaf_fish_per_year.csv")



```


