---
title: "1.1 data wrangling"
author: "TG"
date: "2023-12-10"
output:
   prettydoc::html_pretty:
    theme: cayman
    highlight: github
---


```{r message=TRUE, warning=TRUE, include=FALSE}

library(tidyverse)
library(ggplot2)
library(stringr)
library(Hmisc)
library(rfishbase)
library(lubridate)

select <- dplyr::select
rename <-dplyr::rename
```

# Fish data wrangling

load data

```{r}
# clean and orgenise data from 2015-2021

load("G:/My Drive/PHD/Ecocamp-2022/data/until 2021/fish_full_data.Rdata") # old data

# new fish data from fall 2022

EcoCamp_2022_raw <- read_csv("data/2022/EcoCamp 2022.csv")

data_2022<-EcoCamp_2022_raw
```


```{r}

# change firsy/second observer to the observer name

data_2022$Observer <- ifelse(data_2022$Observer == "first", 
                            data_2022$`First Observer.x`,
                            data_2022$`Second Observer.x`)

# keep only relevant column

data_2022 <- data_2022 %>%
  rename(Site = Location, Abundance = Amount) %>%
  mutate(Date = ymd(Date),Year = year(Date)) %>% 
  select(Site,Knoll,Year,Date,Observer,Species, Abundance, Length,Confidence)

# convert column types 

data_2022$Site = as.character(data_2022$Site) 
data_2022$Year = as.character(data_2022$Year) 
data_2022$Knoll = as.character(data_2022$Knoll)
# change site name to NR

data_2022$Site[data_2022$Site == "Reserve"] = "NR"

# In cases where there is no confidance convert to 0 

data_2022$Confidence[is.na(data_2022$Confidence)]<-0

```


Insert correction to knolls number

knoll 21 in Princess should change to 22 after the storm (2020)
knoll 10 in the Princess should change to 23 after the storm (2020)

Those knoll was documented under the same knoll but the knoll was destroy in the storm and we actually surveyed different knoll!


```{r}

data_2022$Knoll<-ifelse(data_2022$Site =="Princess" & data_2022$Knoll =="21",22,data_2022$Knoll)

data_2022$Knoll<-ifelse(data_2022$Site =="Princess" & data_2022$Knoll =="10",23,data_2022$Knoll)

```


check species names to match fish full data and the updated name accurding to fishbase


```{r}
# list of distinct species

sp_2022<-data_2022 %>% distinct(Species)

# check which fish are valdiate accurding to fishbase

valid_name <-validate_names(sp_2022$Species)

sp_2022$valid<-ifelse(sp_2022$Species %in% valid_name,"valid","not valid")

# remove valid species and sp/spp observations

species_v<-sp_2022 %>% filter(sp_2022$valid == "not valid")
species_v<-species_v %>% filter(str_detect(species_v$Species,"spp")==F)
species_v<-species_v %>% filter(str_detect(species_v$Species,"sp.")==F)
 
# match to the new and accepted name

new_name<-species_v$Species %>% synonyms() %>% 
 select(provided_name = synonym, valid_name = Species, Comment = Status) %>% 
  filter(Comment == "synonym") %>% rename(Species = provided_name)

# join and replace the previos name

data_2022<-left_join(data_2022,new_name,by="Species")

data_2022$Species<-ifelse(!is.na(data_2022$valid_name),
                                data_2022$valid_name,
                                data_2022$Species)

data_2022<-data_2022[,1:9]

```

change species that are found in diffrent distribtion range to the local species

```{r}
data_2022$Species[data_2022$Species == "Dascyllus aruanus"]<-"Dascyllus abudafur"
data_2022$Species[data_2022$Species == "Macropharyngodon bipartitus"]<-"Macropharyngodon marisrubri"
data_2022$Species[data_2022$Species == "Aetobatus narinari"]<-"Aetobatus ocellatus"
data_2022$Species[data_2022$Species == "Amblyglyphidodon leucogaster"]<-"Amblyglyphidodon indicus"
data_2022$Species[data_2022$Species == "Parapriacanthus ransonneti"]<-"Parapriacanthus guentheri"
data_2022$Species[data_2022$Species == "Eviota sebreei"]<-"Eviota punyit"
data_2022$Species[data_2022$Species == "Eviota prasites"]<-"Eviota zebrina"
data_2022$Species[data_2022$Species == "Canthigaster coronata"]<-"Canthigaster cyanospilota"
data_2022$Species[data_2022$Species == "Hemigymnus fasciatus"]<-"Hemigymnus sexfasciatus"

```

check which species are new compared to the data collected up to 2021

```{r}

sp_2022<-data_2022 %>% distinct(Species)

sp_all<-fish_full_data%>% distinct(Species,.keep_all = T) %>% select(Species,Species_2015,Family,Status,a,b)

sp_2022$Species[which(!sp_2022$Species %in% sp_all$Species)]

```

* `Pycnochromis dimidiatus` - updated name, need to change in the full data
* `Stegastes lacrymatus` - updated name, need to change in the full data
* `Gymnothorax johnsoni`,`Gymnomuraena zebra` and `Caesio suevica` - new species in the data


change the updated names in fish_full_data

```{r}

fish_full_data$Species[fish_full_data$Species == "Chromis dimidiata"]<-"Pycnochromis dimidiatus"
fish_full_data$Species[fish_full_data$Species == "Plectroglyphidodon lacrymatus"]<-"Stegastes lacrymatus"



```

add family, coefficient and transient/reef associated status to the species


```{r}
sp_all<-fish_full_data%>% distinct(Species,.keep_all = T) %>% select(Species,Species_2015,Family,Status,a,b)

sp_2022<-left_join(sp_2022,sp_all)

```

Fill in the missing data for the 3 species

*I'll add the a and b later (when fishbase will work)*

```{r}
sp_2022$Species_2015<-ifelse(sp_2022$Species == "Gymnothorax johnsoni"|
                               sp_2022$Species == "Gymnomuraena zebra",
                             "Muraenidae",
                             sp_2022$Species_2015)



sp_2022$Family<-ifelse(sp_2022$Species == "Gymnothorax johnsoni"|
                         sp_2022$Species == "Gymnomuraena zebra",
                             "Muraenidae",
                             sp_2022$Family)


sp_2022$Status<-ifelse(sp_2022$Species == "Gymnothorax johnsoni"|
                         sp_2022$Species == "Gymnomuraena zebra",
                             "C",
                             sp_2022$Status)



sp_2022$Species_2015<-ifelse(sp_2022$Species == "Caesio suevica",
                             "Caesio spp.",
                             sp_2022$Species_2015)



sp_2022$Family<-ifelse(sp_2022$Species == "Caesio suevica",
                             "Caesionidae",
                             sp_2022$Family)


sp_2022$Status<-ifelse(sp_2022$Species == "Caesio suevica",
                             "T",
                             sp_2022$Status)


```


```{r}
data_2022<-left_join(data_2022,sp_2022)
```

Add `survey_id` and `year_month` columns

```{r}

data_2022<-data_2022 %>%  mutate("survey_id"= paste(Site,Knoll,Date,sep = "_"),.after = Date)

data_2022 <- data_2022 %>%  mutate("year_month"= paste(data_2022$Year,"b",sep=" "),.after = Date)

data_2022 <- data_2022 %>% relocate(9:11,.after = Family)

```

remove new surveyor data:


```{r}
unique(data_2022$Observer)

to_remove<-c("Yuval Goth","Inbar Dahan","Dana Kruger","Roi Holzman")

data_2022 <- data_2022 %>% filter(!Observer %in% to_remove)

```


# for assaf


```{r}

old_data<-fish_full_data %>% select(colnames(data_2022))

assaf_fish<-bind_rows(old_data,data_2022)

one_observer<-list()
loop<-1
set.seed(1)

for (i in unique(assaf_fish$survey_id)) {
  
  data <- assaf_fish %>% filter(survey_id == i)
  
  obs_amount <- length(unique(data$Observer))
  obs_name <- unique(data$Observer)
  chosen_obs <- sample(obs_name,1)  
 
  filter_data<-data %>% filter(Observer == chosen_obs)
  one_observer[[loop]]<-filter_data
  loop<-loop+1
  
  
}

assaf_fish<- bind_rows(one_observer)

rm(data,filter_data)
```




```{r}
assaf_fish$Year<-as.numeric(assaf_fish$Year)


assaf_fish <- assaf_fish %>%
  filter(Year > 2017,Confidence <1) %>%
  filter(! Knoll %in% c("Tamar",
                    "South bridge reserve",
                    "south bridge reserve",
                    "North bridge reserve",
                    "north bridge reserve"))

assaf_fish_per_knoll<-assaf_fish %>%
  group_by(Site,Knoll,year_month,Species) %>%
  summarise("Abundance" = sum(Abundance))


assaf_fish_per_site<-assaf_fish %>%
  group_by(Site,year_month,Species) %>%
  summarise("Abundance" = sum(Abundance))

assaf_fish_per_year <-assaf_fish %>%
  group_by(year_month,Species) %>%
  summarise("Abundance" = sum(Abundance))


write.csv(assaf_fish,"assaf_fish.csv")
write.csv(assaf_fish_per_knoll,"assaf_fish_per_knoll.csv")
write.csv(assaf_fish_per_site,"assaf_fish_per_site.csv")
write.csv(assaf_fish_per_year,"assaf_fish_per_year.csv")



```


